# Exercise 8 solution

```{r setup, include = FALSE}
pacman::p_load(tidyverse)

obr_data <- read_csv("saved_data/obr_data.csv")
```

## Question 
Choose an appropriate visualisation to investigate the change in employment rate between 2008 and 2024. Generate this visualisation and comment on your findings.

### Solution {.unnumbered}
A **line graph** would generally be the most appropriate visualisation to show a numeric variable over time. The visual marking on the graph would be a line, defined by time on the x-axis and employment rate on the y-axis. 

To create this line graph, we must ensure that all information needed is included in the data. In this case, we will need to create a new `time` variable, beginning at 1, that combines year and quarter to make up the x-axis:

```{r create time variable}
obr_data <- mutate(obr_data,
                   time = (year - 2008) * 4 + quarter)
```

To build the line graph, we first use the `ggplot` function and specify the data that we wish to display on the coordinate system. We then use the `geom_line` function to visualise the data, specifying the variables thatdefine this visualisation:

```{r line graph employment rate}
# Generate the chart area and specify the data
ggplot(data = obr_data) +
  # Add a line, defined by time on the x-axis, and employment rate on the y
  geom_line(aes(x = time, y = employment_rate))
```
