---
code-annotations: hover
---

# Exercise 4 solutions

```{r setup, include = FALSE}
pacman::p_load(tidyverse, haven, Hmisc)

ehs_tidy <- read_csv("saved_data/ehs_tidy.csv") %>% 
  mutate(across(where(is.character), factor))

```

## Question 1
How many respondents had both weekly rent and mortgage payments given? What are the potential reasons for this?

### Solution {.unnumbered}
We can combine the `filter` and `count` functions to answer the first part of this question:

```{r responders with both rent and mortgage}
ehs_tidy %>% 
  filter(!is.na(weekly_rent), !is.na(weekly_mortgage)) %>% 
  count()
```

There were 105 respondents with both weekly rent and mortgage payments. 

To understand why this is, we could first view this data (or a summary of the data) to see what other characteristics these respondents share:

```{r view rent and mortgage payees}
ehs_tidy %>% 
  filter(!is.na(weekly_rent), !is.na(weekly_mortgage)) %>% 
  summary()
```

All respondents in this group owned and lived in their own home. Most were leasehold properties, suggesting some of the weekly rent refers to lease payments. Other potential reasons could include shared ownership (which is not given as an option for tenure type), or respondents that lived with renters in the same property.

## Question 2
Combine the weekly rent and mortgage variables into a single weekly payment variable.

### Solution {.unnumbered}
Where only one value has been recorded, we want to use this in the new variable. Where both have been recorded, we will need to add the values together to get a weekly total. 

There are a few different ways to do this. The first is to include an `if_else` statement in the `mutate` function, changing how the variable is calculated whether either value is missing or not:

```{r weekly total if_else}
ehs_tidy_ex4 <- ehs_tidy %>% 
  mutate(weekly_total = if_else(is.na(weekly_rent) | #<1>
                                  is.na(weekly_mortgage), #<1>
                                coalesce(weekly_rent, weekly_mortgage), #<2>
                                weekly_rent + weekly_mortgage)) #<3>
```

1. If either `weekly_rent` or `weekly_mortgage` are missing
2. Then return the non-missing value
3. Or else (if both are NOT missing), return the sum of these values

## Question 3
Create a summary table containing the mean, median, standard deviation, and the upper and lower quartiles of the weekly payment (rent and mortgage combined) for each region. What, if anything, can you infer about the distribution of this variable based on the table?

### Solution {.unnumbered}
```{r summarise total payments}
ehs_tidy_ex4 %>% 
  group_by(region) %>% #<1>
  summarise(mean_payment = wtd.mean(weekly_total, weights = weighting, #<2>
                                    na.rm = TRUE), #<2>
            median_payment = wtd.quantile(weekly_total, #<3>
                                          weights = weighting, #<3>
                                          probs = .5, na.rm = TRUE), #<3>
            sd_payment = sqrt(wtd.var(weekly_total, weights = weighting, #<4>
                                      na.rm = TRUE)), #<4>
            lq_payment = wtd.quantile(weekly_total, weights = weighting, #<5>
                                      probs = .25, na.rm = TRUE), #<5>
            uq_payment = quantile(weekly_total, weights = weighting, #<6>
                                  probs = .75, na.rm = TRUE)) %>% #<6>
  ungroup() #<7> 
```

1. Calculate summaries per region
2. Return the weighted mean
3. Return the weighted median (the 50th percentile)
4. Return the weighted standard deviation (the square root of the weighted variance)
5. Return the weighted lower quartile (the 25th percentile)
6. Return the weighted upper quartile (the 75th percentile)
7. Don't forget to `ungroup`

There are big differences between most mean and medians across regions, indicating that the data are not normally distributed. If we use the approximate 95% range formula (mean $\pm$ (2 $\times$ sd)), we would get negative values for almost all regions. Negative payments do not make sense in this context, confirming that the data are not normally distributed. 

In this case, the median and IQR should be give, not the mean and standard deviation.
